Reanalysis Step 2: Multiple Imputation

### 2025-08-17 reanalysis-2-multiple_imputation.qmd
set mirror & seed
parameterize test/final, print out analysius model, consistent seeds

```{r setup}
#| include: false
# Setup analysis environment
source("dependencies.R")
setup_analysis()

# Additional packages for imputation
library(mice)
library(VIM)
library(naniar)
```

@@

# Select only imputation variables
df_mice <- df_for_imputation |>
  dplyr::select(dplyr::all_of(imputation_vars))
```

@@

@@

method_table |>
  kable(caption = "Default imputation methods by variable")

@@

# Don't impute these (already complete)
custom_methods["e_mrn_deidentified"] <- ""
custom_methods["ever_lapse_binary"] <- ""
custom_methods["glaucoma_bef_hitplus_cat"] <- ""  # Already recoded
custom_methods["otherretina_bef_hitplus_cat"] <- ""  # Already recoded

@@

# Ensure exposure is used as predictor
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0
```

@@

```{r run-imputation}
# Set seed for reproducibility
set.seed(2024)

# Number of imputations (typically 20-40 for small amounts of missingness)
n_imputations <- 30

# Run mice imputation
# Note: This may take several minutes depending on dataset size
message("Starting multiple imputation with ", n_imputations, " imputations...")
message("This may take several minutes...")

mice_results <- mice(
  df_mice,
  m = n_imputations,
  method = custom_methods,
  predictorMatrix = pred_matrix,
  maxit = 20,  # Number of iterations per imputation
  seed = 2024,
  printFlag = TRUE
)

@@

# Person DR - compare observed vs imputed
if(sum(is.na(df_mice$person_dr)) > 0) {
  # Check distribution of imputed values
  prop_table_dr <- with(mice_results, table(person_dr)) |> 
    lapply(prop.table) |> 
    do.call(rbind, args = _)

  message("Proportion of DR severity (0=No DR, 1=NPDR, 2=PDR) across imputations:")
  print(summary(prop_table_dr))
}

# Check imputation quality for categorical variables
# Cataract surgery
if(sum(is.na(df_mice$catsurg_before_hitplus_cat)) > 0) {
  prop_table <- with(mice_results, table(catsurg_before_hitplus_cat)) |> 
    lapply(prop.table) |> 
    do.call(rbind, args = _)

  message("Proportion of cataract surgery across imputations:")
  print(summary(prop_table))
}
```

>>>>

```{r setup}
#| include: false
# Setup analysis environment (includes seed setting)
source("dependencies.R")
setup_analysis(seed = 2025)

# Additional packages for imputation
library(mice)
library(VIM)
library(naniar)

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 5 iterations (fast)
# final = 30 imputations, 20 iterations (slow but thorough)
analysis_mode <- "test"  # Change to "final" for production run

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 5
  message("Running in TEST mode: ", n_imputations, " imputations, ", n_iterations, " iterations")
} else if (analysis_mode == "final") {
  n_imputations <- 30
  n_iterations <- 20
  message("Running in FINAL mode: ", n_imputations, " imputations, ", n_iterations, " iterations")
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

@@

# Select only imputation variables
df_mice <- df_for_imputation |>
  dplyr::select(dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2), 
                      labels = c("No_DR", "NPDR", "PDR"),
                      ordered = TRUE)
  )
```

@@

# Print initial imputation model summary
message("=== IMPUTATION MODEL SETUP ===\n")
message("Total variables in dataset: ", ncol(df_mice))
message("Variables to be imputed: ", sum(mice_dry$nmis > 0))
message("Total missing values: ", sum(mice_dry$nmis))
message("\nMissing data by variable:")
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))

print(miss_vars)

@@

message("\nDefault imputation methods:")
print(method_table)

@@

# Don't impute these variables (but they ARE still used as predictors!)
# Method = "" means: use as predictor but don't impute
custom_methods["e_mrn_deidentified"] <- ""  # ID - never impute
custom_methods["ever_lapse_binary"] <- ""  # Complete - use as predictor only
custom_methods["glaucoma_bef_hitplus_cat"] <- ""  # Already recoded to 0/1
custom_methods["otherretina_bef_hitplus_cat"] <- ""  # Already recoded to 0/1
custom_methods["gender_cat"] <- ""  # Likely complete
custom_methods["other_inject"] <- ""  # Likely complete
custom_methods["anti_VEGF"] <- ""  # Likely complete  
custom_methods["focal_laser_flag"] <- ""  # Likely complete
custom_methods["PRP_flag"] <- ""  # Likely complete

# Note: Variables with method = "" are STILL USED as predictors for other variables
# They just aren't imputed themselves

@@

# Ensure exposure is used as predictor
pred_matrix[, "ever_lapse_binary"] <- 1
pred_matrix["ever_lapse_binary", "ever_lapse_binary"] <- 0

# Print final imputation model configuration
message("\n=== FINAL IMPUTATION MODEL ===\n")

# Summary of methods
final_methods <- data.frame(
  Variable = names(custom_methods),
  Method = custom_methods,
  Type = case_when(
    custom_methods == "" ~ "Predictor only (not imputed)",
    custom_methods == "pmm" ~ "Predictive mean matching",
    custom_methods == "polr" ~ "Proportional odds regression",
    custom_methods == "logreg" ~ "Logistic regression",
    custom_methods == "polyreg" ~ "Polytomous regression",
    TRUE ~ custom_methods
  )
) |>
  arrange(Method != "", Variable)

message("Imputation methods by variable:")
print(final_methods, n = 50)

# Show which variables will be imputed
vars_to_impute <- names(custom_methods)[custom_methods != ""]
message("\nVariables that WILL be imputed (", length(vars_to_impute), " total):")
print(vars_to_impute)

# Show predictor matrix summary
message("\nPredictor matrix summary:")
message("- Each variable uses ", sum(pred_matrix[1,]), " predictors on average")
message("- Variables used as predictors:")
predictors <- names(which(colSums(pred_matrix) > 0))
print(predictors)

# Print model parameters
message("\n=== IMPUTATION PARAMETERS ===")
message("Mode: ", analysis_mode)
message("Number of imputations (m): ", n_imputations)
message("Max iterations per imputation (maxit): ", n_iterations)
message("Random seed: 2025")
```

@@

```{r run-imputation}
# Set seed for reproducibility
set.seed(2025)

# Run mice imputation
# Note: Parameters are set based on analysis_mode in setup
message("Starting multiple imputation...")
message("Mode: ", analysis_mode)
message("Imputations: ", n_imputations)
message("Iterations: ", n_iterations)

if (analysis_mode == "test") {
  message("This should take a few minutes...")
} else {
  message("This may take 15-30 minutes or more...")
}

mice_results <- mice(
  df_mice,
  m = n_imputations,
  method = custom_methods,
  predictorMatrix = pred_matrix,
  maxit = n_iterations,  # Uses n_iterations from setup
  seed = 2025,
  printFlag = TRUE
)

@@

# Person DR - compare observed vs imputed (categorical)
if(sum(is.na(df_mice$person_dr)) > 0) {
  # For categorical variables, check proportions across imputations
  # Extract each imputation and tabulate
  dr_props <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    prop.table(table(imp_data$person_dr))
  })

  message("Proportion of DR severity across imputations:")
  message("No_DR: ", round(mean(dr_props[1,]), 3), " (SD: ", round(sd(dr_props[1,]), 3), ")")
  message("NPDR: ", round(mean(dr_props[2,]), 3), " (SD: ", round(sd(dr_props[2,]), 3), ")")
  message("PDR: ", round(mean(dr_props[3,]), 3), " (SD: ", round(sd(dr_props[3,]), 3), ")")

  # Visual check if stripplot works with factors
  tryCatch({
    stripplot(mice_results, person_dr ~ .imp, pch = 20, cex = 1.2) |> print()
  }, error = function(e) {
    message("Note: Stripplot not available for categorical person_dr")
  })
}

# Check imputation quality for categorical variables
# Cataract surgery
if(sum(is.na(df_mice$catsurg_before_hitplus_cat)) > 0) {
  # Extract proportions for each imputation
  cat_props <- sapply(1:n_imputations, function(i) {
    imp_data <- complete(mice_results, i)
    prop.table(table(imp_data$catsurg_before_hitplus_cat))
  })

  message("Proportion of cataract surgery across imputations:")
  if(is.matrix(cat_props)) {
    for(i in 1:nrow(cat_props)) {
      message("Level ", rownames(cat_props)[i], ": ", 
              round(mean(cat_props[i,]), 3), " (SD: ", round(sd(cat_props[i,]), 3), ")")
    }
  } else {
    message("Prop = ", round(mean(cat_props), 3), " (SD: ", round(sd(cat_props), 3), ")")
  }
}
```
